Fail Faster

### 2025-03-17 paper/paper.tex
I'm committing.

While neither of these microbenchmarks exhibit this behavior, more complex
generators suffer further performance penalties due to \emph{closure
allocation}. In cases where the compiler cannot statically eliminate it \jwc{``it'' = ``the allocation''},
running a monadic bind allocates a short-lived closure: we allocate a closure
for the continuation, and then immediately jump into it.
In strict functional languages like OCaml and Scala,
Each individual closure allocation is relatively cheap, but doing lots of allocation in a generation
loop is very costly because each allocation brings us closer to the next costly GC pause \cn{}.

\subsubsection{Overhead of Choice Combinators}

@@

\jwc{In practice, the possible options are almost always statically known: in the binary tree generator from earlier, we can tell from the text of the code what
the weights are, what the generators are, and how many there are.
Put more simply, you basically always call weighted union with an \emph{explicit list}, which is then immediately traversed by the weighted union. For this reason, you should not have to incur the cost of allocating this list at run time.
However, almost no compilers (GHC is a notable exception) can figure this out and eliminate the list \cn (list fusion).
}

>>>>

While neither of these microbenchmarks exhibit this behavior, more complex
generators suffer further performance penalties due to \emph{closure
allocation}. In cases where the compiler cannot statically eliminate it \jwc{``it'' = ``the allocation''},
running a monadic bind allocates a short-lived closure: we allocate a closure
for the continuation, and then immediately jump into it.
In strict functional languages like OCaml and Scala,
Each individual closure allocation is relatively cheap, but doing lots of allocation in a
generator is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
Similarly to \jwc{the previous one} this effect is magnified in recursive generators: each iteration
through the recursive loop re-allocates closures for \texttt{bind}s, so the amount of allocation
\emph{per generated value} scales linearly with the number of recursive generator calls.

\subsubsection{Overhead of Choice Combinators}
The monad functions \texttt{return} and \texttt{bind} aren't the only combinators that incur an
abstraction overhead: all of them do! \jwc{I'm so bad at starting new sections...} ``Choice'' combinators like \texttt{weighted\_union}, however
are particularly costly, because they unavoidably allocate a lot.

@@

In practice, \texttt{weighted\_union}s is almost always\footnote{
There are some generators where \texttt{weighted\_union} is passed a list which
was itself the result of a generator: the well-typed STLC term generator used in Section~\ref{section:eval} is an example of this.
} passed an explicitly constructed list, like
\texttt{weighted\_union [(3,g); (1,g'); (2,g'')]}. 
This is because the most common use case for \texttt{weighted\_union} is
to choose between one of the different constructors of an algebraic datatype,
the options for which are always known.
The list passed to \texttt{weighted\_union} allocated at the call site, and then \emph{immediately}
destructed by the function. In principle, since the elements of the list and its length $n$ are known statically,
a compiler could unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function to avoid allocating the list.
In practice however almost~\jwc{Once again, GHC is a notable exception here \cn{}} no compilers perform this kind of optimization.
